Exploration of the MIRT solutions

### 2024-01-16 notebooks/Mirt_solution_exploration.Rmd
Checks `{mirt}` solution, including factor scores and multidimensional parameters

@@

However, the MDISC and MID parameters are possibly estimated from the original,
unrotated solution.

>>>>

source('R/Mirt_toolbox.R', encoding = 'UTF-8')

@@

However, the MDISC and MID parameters are possibly estimated from the original,
unrotated solution. Note that the values of these multidimensional parameters
is invariant to the rotation.

# Model estimation

We start by fitting a model with sample data bundled with the `{mirt}` package
itself. Let's start by using the first multidimensional example:

```{r fit-model}
data <- expand.table(LSAT7)
itemstats(data)

mod2 <- mirt(data, 2)
coef(mod2)
summary(mod2, rotate = 'oblimin') #oblimin rotation
```

## Parameter computation

The default model is a 2-PL, so the items only have two discrimination
parameters (one for each dimension), and an intercept parameter. Parameters `g`
and `u` are restricted to `0` and `1`, respectively.

After applying the `oblimin` rotation, the item parameters and covariances are:

```{r rotated-irt-params}
coefs <- coef(mod2, rotate = 'oblimin') |> map(as_tibble)

item_pars <- coefs |>
  bind_rows()      |>
  slice(1:5)       |>
  select(a1:d)     |>
  rownames_to_column(var = "item")

cov_pars <- coefs[['GroupPars']] |>
  pivot_longer(everything())     |>
  deframe()

cov_matrix <- cov_pars[c('COV_11', 'COV_21', 'COV_21', 'COV_22')] |>
  matrix(nrow = 2)
```

In order to check the result expressed in factor-analytic metric, we turn the
factor-analytic parameters into IRT metric to see whether they are coincident:

```{r param-equivalence}
sum <- summary(mod2, rotate = 'oblimin') #oblimin rotation
sum$rotF |> as_tibble() |> mutate(
  across(everything(), ~1.702 * . / sqrt(1-as.numeric(sum$h2)))
)
```

The equivalence holds.

## Factor scores

Factor scores are computed by means of the `fscores()` function. By default,
`{mirt}` applies an `oblimin` rotation.

We use the original LSAT7 response patterns to compute the factor scores:

```{r compute-fscores}
scores <- fscores(mod2, response.pattern = LSAT7[, 1:5])
```

## Multidimensional IRT parameters

We compute the multidimensional IRT parameters using the `{mirt}` functions, and
also the IRT parameters of the rotated solution, and compare them.

```{r multidim-params}
mod2 |> MDISC()
mod2 |> MDIFF()

item_pars |>
  compute_mirt_params(d, a1:a2, cov_matrix = cov_matrix, dir_out = 'cos')
```

We see that they have the same values. Also, we can use the original, unrotated
solution:

```{r multidim-params-unrotated}
coefs_unrot <- coef(mod2) |> map(as_tibble)

item_pars_unrot <- coefs_unrot |>
  bind_rows()                  |>
  slice(1:5)                   |>
  select(a1:d)                 |>
  rownames_to_column(var = "item")

item_pars_unrot |>
  compute_mirt_params(d, a1:a2, dir_out = 'cos')
```

As expected, the MDISC and the location component of the MDIFF are coincident,
but the direction cosines change.

A visual inspection of the `MDISC()` and `MDIFF()` functions shows that the
parameters are computed applying Reckase's formulae to the unrotated solution.
